Red Wines Exploration by Natalia Lokshina
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(PerformanceAnalytics)
library(RColorBrewer)
```

This report explores a dataset containing properties of almost 1,600 red wines.

# Univariate Plots Section

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv("wineQualityReds.csv")
```

First we need to look at some summaries of the data in order to get the idea
of the data structure.

```{r echo=FALSE, Data_Structure}
str(df)
```

```{r echo=FALSE, Data_Summary}
summary(df)
```

Our dataset contains 1,599 observations of red wines with 11 variables
on the chemical properties of the wine and 1 variable on quality.
At least 3 wine experts rated the quality of each wine, providing a rating
between 0 (very bad) and 10 (very excellent).

Lets look at the distribution of the individual variables.

```{r echo=FALSE, Univariate_Plots}
# quality distribution
ggplot(aes(x = quality), data = df) +
  geom_bar() +
  scale_x_continuous(breaks=0:10)
```

From this bar plot we can see, that more than half of the wines in our dataset
got quality rating 5 or 6. None of the wines were rated lower than 3 and higher
than 8.

```{r echo=FALSE}
# fixed acidity and volatile acidity distributions
p1 <- ggplot(aes(x = fixed.acidity), data = df) +
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(breaks = seq(0, 16, 1))

p2 <- ggplot(aes(x = volatile.acidity), data = df) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(breaks = seq(0, 2, 0.2))

grid.arrange(p1, p2, ncol = 2)
```

The distributions of the fixed acidity and volatile acidity are right skewed.
They both have a noticeable pick: fixed acidity somewhere between 6.5 and 8.5
and volatile acidity between 0.3 and 0.7.
Since the distributions look similar, it will be interesting to see if these
variables are correlated.

```{r echo=FALSE}
# citric acid distribution
ggplot(aes(x = citric.acid), data = df) +
  geom_histogram(binwidth = 0.05)

```

```{r}
# calculate the amount of values = 1, 0 and < 0.05 (1st bin)
print(dim(df[df$citric.acid == 1,]))
print(dim(df[df$citric.acid == 0,]))
print(dim(df[df$citric.acid < 0.05,]))
```

The citric acid distribution looks multimodal with the biggest pick close to 0
and the other 2 about 0.25 and 0.5.
There is one outlier with the value 1.
Also out of 274 values less then 0.05, which fall into the first bin, almost
half (132) have value 0. I wonder if they represent missing data and not
the actual level of citric acid.

```{r echo=FALSE}
# residual sugar distribution
ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram(binwidth = 0.5)
```

On the residual sugar histogram we can see a few outliers which give
the distribution a big right skew. Let's zoom in.

```{r echo=FALSE}
# residual sugar distribution zoomed in
ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 7)) +
  scale_x_continuous(breaks = seq(0, 7, 0.5))
```

```{r echo=FALSE}
# statistics for residual sugar
summary(df$residual.sugar)
```

From the histogram and the statistics we can tell that half of the values are
between 1.9 and 2.6. Interesting to note that the mean (2.539) is very close
to 3rd quartile (2.6), which explains a long right skew.

```{r echo=FALSE}
# chlorides distribution
p1 <- ggplot(aes(x = chlorides), data = df) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(0, 0.7, 0.1))

p2 <- ggplot(aes(x = chlorides), data = df) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(0, 0.15, 0.025)) +
  coord_cartesian(xlim = c(0, 0.15))

grid.arrange(p1, p2, ncol = 2)
```

```{r echo=FALSE}
# chlorides statistics
summary(df$chlorides)
```

The distribution of chlorides is similar to residual sugar.
50% of values are between 0.07 and 0.09.

```{r echo=FALSE}
# free sulfur dioxide and total sulfur dioxide distributions
p1 <- ggplot(aes(x = free.sulfur.dioxide), data = df) +
  geom_histogram(binwidth = 2)

p2 <- ggplot(aes(x = total.sulfur.dioxide), data = df) +
  geom_histogram(binwidth = 10)

grid.arrange(p1, p2, ncol = 2)
```


```{r echo=FALSE}
# free sulfur dioxide statistics
summary(df$free.sulfur.dioxide)
```

```{r echo=FALSE}
# free total dioxide statistics
summary(df$total.sulfur.dioxide)
```

The distributions of free sulfur dioxide and total sulfur dioxide also look
very similar. They are both right skewed.
The free sulfur dioxide values lie between 1 and 72 with half of them
between 7 and 21 and median 14.
The total sulfur dioxide values have minimum of 6 and maximum of 289.
There are few outliers close to 289, but 75% of the values are less than 62.
I would expect these two variables to be correlated.

```{r echo=FALSE}
# density distribution
ggplot(aes(x = density), data = df) +
  geom_histogram(binwidth = 0.001)
```

```{r echo=FALSE}
# density statistics
summary(df$density)
```

Density looks normally distributed.
The mean and median are almost identical: 0.9967 and0.9968.
Interesting to note that all values lie in a very small range: 0.0136.
Interquartile range is just 0.0022.

```{r echo=FALSE}
# pH distribution
ggplot(aes(x = pH), data = df) +
  geom_histogram(binwidth = 0.05)
```

```{r echo=FALSE}
# pH statistics
summary(df$pH)
```

The distribution of pH is also close to normal with mean = 3.311 and
median = 3.310.

```{r echo=FALSE}
# sulphates distribution
ggplot(aes(x = sulphates), data = df) +
  geom_histogram(binwidth = 0.05)
```

From this histogram we can see that sulphates distribution is right skewed.
There are a few outliers close to 2.

```{r echo=FALSE}
# alcohol distribution
ggplot(aes(x = alcohol), data = df) +
  geom_histogram(binwidth = 0.5)
```

```{r echo=FALSE}
# alcohol statistics
summary(df$alcohol)
```

The alcohol distribution is again right skewed.
The lowest level of alcohol is 8.4 and the highest is 14.9.
Most of the wines have alcohol level somewhere between 9.5 and 11.5.


# Univariate Analysis

### What is the structure of your dataset?

There are 1599 red wine observations with 12 variables:

 - fixed acidity
 - volatile acidity
 - citric acid
 - residual sugar
 - chlorides
 - free sulfur dioxide
 - total sulfur dioxide
 - density
 - pH
 - sulphates
 - alcohol
 - quality

All variables have numeric values.
Quality is the only categorical variable which has values between 0 (very bad)
and 10 (very excellent).

Other observations:

 - The citric acid distribution looks multimodal with the biggest pick
 close to 0 and the other 2 about 0.25 and 0.5.
 - The distributions of the fixed acidity, volatile acidity, residual sugar,
 chlorides, free sulfur dioxide, total sulfur dioxide, sulphates and alcohol
 are right skewed.
 - The distributions of density and pH are normal.
 - The quality levels are between 3 and 8. Most of the wines have quality
 rating of 5 and 6.

### What is/are the main feature(s) of interest in your dataset?

The main feature is quality. I want to see if any of the provided chemical
properties correlated with the quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

So far it is difficult to say which features are going to be correlated with
the quality of wine.

### Did you create any new variables from existing variables in the dataset?

I created a factor variable from quality.
It will help me in bivariate exploration section.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I noticed that out of 274 values of citric acid less then 0.05,
which fall into the first bin, almost half (132) have value 0.
I think they represent missing data and not the actual level of citric acid.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
# drop first column "X" and replace "quality" with factor type
df2 <- df[,-1]

# calculate correlation matrix
corr_matr <- round(cor(df2),3)
corr_matr
```

```{r echo=FALSE}
# plot correlation matrix
chart.Correlation(df2, histogram=TRUE, pch=19)
```

Here we can see that there are no strong correlations between any variables.
Some variables moderately correlated. The strongest relationship is between
fixed acidity and pH (-0.683).
Let's look closer.

```{r echo=FALSE}
# correlation between fixed acidity and volatile acidity
ggplot(df2, aes(x=fixed.acidity, y=volatile.acidity)) + 
  geom_jitter() +
  stat_smooth(method = "lm", color="red")
```

Earlier I was wondering if fixed acidity and volatile acidity are correlated.
We can see now that they have weak negative correlation (-0.256).
But both of them have moderate correlation with citric acid.

```{r echo=FALSE}
# correlation between fixed acidity / volatile acidity and citric acid
p1 <- ggplot(df2, aes(x=fixed.acidity, y=citric.acid)) + 
  geom_jitter() +
  stat_smooth(method = "lm", color="red")
p2 <- ggplot(df2, aes(x=volatile.acidity, y=citric.acid)) + 
  geom_jitter() +
  stat_smooth(method = "lm", color="red")
grid.arrange(p1, p2, ncol = 2)
```

We can see that fixed acidity and citric acid have positive moderate
correlation (0.672), and volatile acidity and citric acid have negative
moderate correlation (-0.552).

```{r echo=FALSE}
# correlation between free sulfur dioxide and total sulfur dioxide
ggplot(df2, aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) + 
  geom_jitter() +
  stat_smooth(method = "lm", color="red")
```

As I expected free sulfur dioxide and total sulfur dioxide have moderate
positive correlation (0.668). The higher level of free sulfur dioxide,
the higher level of total sulfur dioxide.

Now I want to look what has correlation with quality.

```{r echo=FALSE}
# correlations with quality
corr_matr["quality",]
```

According to these data quality is moderately correlated only with volatile
acidity and alcohol. The smallest correlation is with residual sugar.

```{r echo=FALSE}
# correlation between quality and volatile acidity
ggplot(df2, aes(x=quality, y=volatile.acidity)) + 
  geom_jitter(alpha=0.5) +
  scale_x_continuous(breaks=0:10) +
  stat_smooth(method = "lm", color="red")
```

```{r echo=FALSE}
# Create a new column f_quality, factor type of quality column
df2$f_quality <- factor(df2$quality)

# correlation between quality and volatile acidity
ggplot(df2, aes(x=f_quality, y=volatile.acidity)) +
  geom_boxplot()
```

These two plots support the correlation coefficient we got earlier (-0.391).
In boxplot we can also see lack of correlation quality rating 7 and 8.

```{r echo=FALSE}
# correlation between quality and alcohol
ggplot(df2, aes(x=f_quality, y=alcohol)) + 
  geom_jitter(alpha=0.3, color="tomato") +
  geom_boxplot(alpha=0)
```

```{r}
# statistics for alcohol by quality
by(df2$alcohol, df2$quality, summary)
```

Correlation coefficient for alcohol and quality is 0.476. We can see it on
our plot also.
But what is interesting, the median alcohol level for quality rate 3 and 4
is actually higher than for quality rate 5.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There are no strong correlations between any variables.
Some variables moderately correlated.

Fixed acidity and volatile acidity have moderate correlation with citric acid.

Quality is moderately correlated only with volatile acidity and alcohol.

Even though the correlation between alcohol and quality is positive,
the median alcohol level for quality rate 3 and 4 is actually higher
than for quality rate 5.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

As I expected free sulfur dioxide and total sulfur dioxide have moderate
positive correlation.

### What was the strongest relationship you found?

The strongest relationship is between fixed acidity and pH (-0.683).

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
# correlation between fixed acidity, volatile acidity and citric acid
ggplot(df2, aes(x=fixed.acidity, y=volatile.acidity, color=citric.acid)) + 
  geom_jitter()
```

On this scatter plot we can see the same result as earlier.
The citric acid is higher for higher values of fixed acidity and lower values
of volatile acidity.

```{r echo=FALSE}
# correlation between free sulfur dioxide, total sulfur dioxide and quality
ggplot(df2, aes(x=free.sulfur.dioxide,
                y=total.sulfur.dioxide,
                color=f_quality)) + 
  geom_jitter() +
  scale_color_brewer(palette = "Blues")
```

```{r echo=FALSE}
# same faceted by quality
ggplot(df2, aes(x=free.sulfur.dioxide,
                y=total.sulfur.dioxide,
                color=f_quality)) +
  geom_jitter() +
  facet_wrap(~f_quality) +
  stat_smooth(method = "lm")
```

Here we can see lack of correlation between free sulfur dioxide or total sulfur
dioxide and quality. For each quality level there is the same pattern of
relationship between free sulfur dioxide and total sulfur dioxide.

```{r echo=FALSE}
# correlation between fixed acidity, volatile acidity and quality
ggplot(df2, aes(x=fixed.acidity, y=volatile.acidity, color=f_quality)) + 
  geom_jitter() +
  scale_color_brewer(palette = "Blues") +
  theme_dark()
```

On this plot dark blue points are mostly on the bottom part of plot. That means
that there is a negative correlation between volatile acidity and quality,
which we had shown earlier. Also we can see a weak negative correlation between
fixed acidity and volatile acidity.

```{r echo=FALSE}
# same faceted by quality
ggplot(df2, aes(x=fixed.acidity, y=volatile.acidity, color=f_quality)) + 
  geom_jitter() +
  scale_color_brewer(palette = "Blues") +
  facet_wrap(~f_quality) +
  stat_smooth(method = "lm", color="red") +
  theme_dark()
```

Looking at scatter plots of fixed acidity and volatile acidity separately for
each quality level it seems that correlation between them is stronger for
quality rates 7 and 8 than for quality rates 4, 5 and 6.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Observations in this part supported earlier conclusions.
Citric acid is positively correlated with fixed acidity and negatively
correlated with volatile acidity.
There is a negative correlation between volatile acidity and quality

### Were there any interesting or surprising interactions between features?

Fixed acidity and volatile acidity seem to have stronger correlation for
quality rates 7 and 8 than for quality rates 4, 5 and 6.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Quality distribution
ggplot(aes(x = quality), data = df) +
  geom_bar(color = 'black', fill = '#099DD9') +
  scale_x_continuous(breaks=0:10) +
  labs(title ="Quality distribution",
       x = "Quality rating",
       y = "Friquancy")
```

### Description One
From this bar plot we can see, that more than half of the wines in our dataset
got quality rating 5 or 6. None of the wines were rated lower than 3 and higher
than 8. So we don't have very excellent wines or very bad wines in the dataset.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Correlation between Alcohol and Quality
ggplot(df2, aes(x=f_quality, y=alcohol)) + 
  geom_jitter(alpha=0.3, color="tomato") +
  geom_boxplot(alpha=0) +
  labs(title ="Correlation between Alcohol and Quality",
       x = "Quality rating",
       y = "Alcohol (% by volume)")
```

### Description Two
Correlation coefficient for alcohol and quality is 0.476. We can see it on
our plot also.
But what is interesting, the median alcohol level for quality rate 3 and 4
is actually higher than for quality rate 5.

### Plot Three
```{r echo=FALSE, Plot_Three}
# Correlation between Fixed Acidity, Volatile Acidity and Quality
ggplot(df2, aes(x=fixed.acidity, y=volatile.acidity, color=f_quality)) + 
  geom_jitter() +
  scale_color_brewer(palette = "Blues", name="Quality") +
  theme_dark() +
  labs(title ="Correlation between Fixed Acidity, Volatile Acidity and Quality",
       x = "Fixed Acidity (g / dm^3)",
       y = "Volatile Acidity (g / dm^3)")
```

### Description Three
On this plot dark blue points are mostly on the bottom part of plot. That means
that there is a negative correlation between volatile acidity and quality.
Also we can see a weak negative correlation between
fixed acidity and volatile acidity.

------

# Reflection

Our dataset contains 1,599 observations of red wines with 11 variables on the
chemical properties of the wine and 1 variable on quality. At least 3 wine
experts rated the quality of each wine, providing a rating between 0 (very bad)
and 10 (very excellent).
More than half of the wines in our dataset got quality rating 5 or 6.
Surprisingly, none of the wines were rated lower than 3 and higher than 8.
So we don't have very excellent wines or very bad wines in the dataset.
Quality is moderately correlated only with volatile acidity and alcohol.
The smallest correlation is with residual sugar.

I struggled with finding strong correlations between quality of wine and its
chemical properties.
Also I had problem predicting any correlations based on Univariate Analysis.
Maybe it happened because of lack of knowledge in chemistry. In future work I
would research more on that topic.
After plotting a correlation matrix, I quickly found variables with
relationship.
